\documentclass{article}

\usepackage{amsmath, amsthm}
\usepackage{setspace}
\usepackage{microtype, parskip}
\usepackage[comma,sort&compress]{natbib}
\usepackage{lineno}
\usepackage{docmute}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}

\frenchspacing

\doublespacing

\raggedright

\begin{document}
\linenumbers
\modulolinenumbers[2]

%\maketitle

\begin{titlepage}
  \begin{large}
    \textbf{Title:} How do biological traits affect brachiopod taxonomic survival? A hierarchical Bayesian approach.
  \end{large}

  \textbf{Running title:} How do biological traits affect taxonomic survival?

  \textbf{Author:} Peter D Smits, psmits@uchicago.edu, Committee on Evolutionary Biology, University of Chicago

  \textbf{Keywords:} extinction, macroevolution, macroecology, Paleozoic, selection

  \textbf{Word count:} ?
  
  \textbf{Table count:} 0.
 
  \textbf{Figure count:} 5 main text, 3 supplement.

  \textbf{Data archival location:} If accepted, all data and code necessary to duplicate this analysis will be made available on DRYAD.

\end{titlepage}

\input{abstract}

\input{intro}

\input{methods}

\input{results}

\input{disc}

\section*{Acknowledgements}
I would like to thank K. Angielzcyk, M. Foote, P. D. Polly, and R. Ree for helpful discussion and advice. Additionally, thank you A. Miller for the epicontinental versus open-ocean assignments. This entire study would would not have been possible without the Herculean effort of the many contributors to the Paleobiology Database. In particular, I would like to thank J. Alroy, M. Aberhan, D. Bottjer, M. Clapham, F. F\"{u}rsich, N. Heim, A. Hendy, S. Holland, L. Ivany, W. Kiessling, B. Kr\"{o}ger, A. McGowan, T. Olszewski, P. Novack-Gottshall, M. Patzkowsky, M. Uhen, L. Villier, and P. Wager. This work was supported by a NASA Exobiology grant (NNX10AQ446) to A. Miller and M. Foote. This is Paleobiology Database publication XXX.

\clearpage

\bibliographystyle{evolution}
\bibliography{newbib,packages}

\clearpage

\appendix
\input{appendix}

\end{document}
